### 2019-10-27 Manuscript/manuscript.tex
add intermittent stagnant lid text to disscussion and start abstract

\subsection{Introduction}

Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography. There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models. This chapter seeks to provide a concise review of these records.

\subsection{Broad tectonic history overview}

@@

There is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of the types of low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones in the Proterozoic relative to the Phanerozoic \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective there is strong paleomagnetic support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the the Trans-Hudson orogenic cycle, the Grenville orogenic cycle, and the Appalachian orogenic cycle are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also ecologites preserved in the Trans-Hudson orogen that preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga \citep{Weller2017a}. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.

As an additional example, consider the record of Laurentia in the late Mesoproterozoic. As can be seen in Table 2, there are many high-quality paleomagnetic poles between 1110 and 1070 Ma. These poles constrain rapid motion of Laurentia leading up to collisional orogenesis as illustrated in Figure \ref{fig:Grenville_reconstructions} providing strong evidence for differential plate motion at the time.

>>>>

\subsection{Abstract}

Laurentia was a major continent throughout the majority of the Proterozoic and is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography. There is a rich record of Precambrian paleomagnetic poles from Laurentia as well as an extensive geologic history of tectonism that are both key to evaluating and developing paleogeographic models. Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics

\subsection{Introduction and broad tectonic history}

@@

There is strong evidence both in Laurentia's geological and paleomagnetic record for differential plate tectonic motion between 2.2 and 1.8 Ga. The continued history of accretionary orogenesis and the evaluation of Laurentia's pole path in comparison to other continents from 1.8 Ga onward supports the continual operation of plate tectonics throughout the rest of the Proterozoic and Phanerozoic as well. While this evidence fits with the majority of interpretations of the timing of initiation of modern-style plate tectonics (see summary in \citealp{Korenaga2013a}), there continue to be arguments proposing that a stagnant lid persisted through the Mesoproterozoic Era (1.6 to 1.0 Ga) and into the Neoproterozoic  with plate tectonics not initiating until ca. 0.8 Ga \citep{Hamilton2011a, Stern2018a}. These arguments rest largely on the relative lack of Proterozoic low-temperature high-pressure metamorphic rocks such as blueschists that form in subduction zones \citep{Stern2013a}. An alternative interpretation for this lack of blueschists in the Proterozoic is that such a shift in metamorphic regime is the predicted result of secular evolution of mantle chemistry rather than a harbinger of the onset of plate tectonics \citep{Palin2015a}. While this line of evidence is intriguing, to argue that there was not differential plate tectonic motion in the Paleoproterozoic and Mesoproterozoic is to ignore a vast breadth and depth of geological and paleomagnetic data. From a paleomagnetic perspective, there is strong support for independent and differential motion of the Slave and Superior provinces as is illustrated in Figure \ref{fig:Superior_Slave_recons}. From a geological perspective, the Trans-Hudson orogenic cycle, the Grenville orogenic cycle, and the Appalachian orogenic cycle are all well-explained with a mobilistic interpretation that includes phases of accretionary followed by collisional orogenesis (Fig. \ref{fig:tectonic_history}). One could counter that this perspective results from a plate-tectonic-centric viewpoint that lacks creativity to see the record as resulting from other processes than modern-style plate tectonics. However, in addition to the broad geological record showing an amalgamation of terranes as would be expected to arise through plate tectonics, there are also ecologites preserved in the Trans-Hudson orogen that preserve evidence for high-pressure/low-temperature metamorphic conditions ca. 1.8 Ga \citep{Weller2017a}. Similar to the Himalayan orogen, these rocks are interpreted to be the result of deep continental subduction and exhumation associated with convergent plate tectonics \citep{Weller2017a}. Outside of Laurentia, there are examples of ecologites with geochemical affinity to oceanic crust such as that documented in the ca. 1.9 Ga Ubendian Belt of the Congo craton \citep{Boniface2012a}.

Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}. In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion. In particular, it was been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}. The accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). An additional constraint comes from the presence of reliable paleomagnetic poles themselves that are complete with baked contact tests. In a stagnant lid regime, there would not be sufficient heat flow to sustain a geodynamo and Earth should not have had a geomagnetic field that would have enable the preservation of paleomagnetic poles with positive field tests. The record of these poles also show that there was motion of Laurentia (Figs. \ref{fig:Laurentia_paleolatitude} and \ref{fig:Laurentia_reconstructions}). Using data from Laurentia alone, however, it is difficult to ascertain whether this motion is due to plate tectonic motion or rotation of the entire solid Earth through true polar wander. True polar wander can lead to changing position relative to the spin axis even with a stagnant lid. One interval when the Laurentian paleomagnetic record demands that some of the motion is through differential plate tectonics is in the latest Mesoproterozoic. At that time, the pole path is very well-resolved with many high-quality paleomagnetic poles between 1110 and 1070 Ma (Table 2; \ref{fig:Laurentia_poles}). The progression of the poles requires rotation about an Euler pole that is distinct from a great circle path which would result if the motion were solely due to true polar wander \citep{Swanson-Hysell2019a}. These poles constrain rapid motion of Laurentia leading up to collisional orogenesis associated with the Grenvillian orogeny as illustrated in Figure \ref{fig:Grenville_reconstructions}. These data provide strong evidence for differential plate motion at the time and are inconsistent with a stagnant lid. Rather the orogenic cycle of the Mesoproterozoic bears similarity with that of the Paleozoic and reveals Laurentia to have been a central player in the building of amalgamated continents associated with Rodinia and Pangea.
